---
title: "Chapter 9- Advanced R Programming"
output: html_document
date: "2024-05-18"
---

[click here](http://adv-r.had.co.nz/) to open a book that talks more about advanced programming

# Working with vectors and Vectorizing Functions

-   Previously we've dealt with expressions on single values(Scalar)

-   But R doesn't have scalar values and treats all expressions as vectors, be it single or many expressions

-   This expressions can be referred to as vectorized expressions

-   They perform element-wise operations on data structures

-   Consider code below

```{r}
(x<-2/3)
(y<-x**2)
```

-   Our expressions are working on single values

```{r}
is.vector(2)
is.vector("a")
```

-   The chunk above proves that R takes all single expressions as vectors

-   consider the chunk:

```{r}
(x<-1:4/3)
(y<-x**2)
```

-   R works on vectors using 2 rules:

    1.  Operations are done element-wise
    2.  Vectors are repeated as needed (Recycling process)

    **Element wise operations**

-   The expression `x+y` implies creation of new vectors whose values are the element wise summation of vectors x and y

```{r}
x<-1:5
y<-6:10
x+y
```

-   This can also be written as :

```{r}
(z<-x+y)
```

-   Which can also be written as :

```{r}
z<-vector(length=length(x))
for(i in seq_along(x)){
  z[i]<-x[i]+y[i]
}
z
```

-   The chunk is a breakdown of what R simply does when performing element wise computation for vectors, in our case element wise summation

-   This is the case for all arithmetic expressions and logical expressions involving `|` and `&`

-   The same case also applies form most functions one calls

```{r}
sqrt((1:5)^2)
sin(sqrt((1:5)^2))
```

**Recycling Process**

-   It occurs when you have one vector that is longer than the other

-   In such scenario, R repeats the shorter vector to make it of same length as the long one

-   **NB:** Length of long vector should be divisible by length of shorter vector

-   If this is not the case a warning is displayed but the recycling process is performed

-   Once this is done R can perform operations on the vectors element-wise

```{r}
x<-1:10
y<-1:2
x+y
```

-   The table below shows the recycling process and the element wise performance

|     |     |     |     |     |     |     |     |     |     |     |
|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| x   | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  |
| y   | 1   | 2   | 1   | 2   | 1   | 2   | 1   | 2   | 1   | 2   |
| x+y | 2   | 4   | 4   | 6   | 6   | 8   | 8   | 10  | 10  | 12  |

```{r}
z<-1:3
x+z
```

|     |     |     |     |     |     |     |     |     |     |     |
|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| x   | 1   | 2   | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  |
| z   | 1   | 2   | 3   | 1   | 2   | 3   | 1   | 2   | 3   | 1   |
| x+z | 2   | 4   | 6   | 5   | 7   | 9   | 8   | 10  | 12  | 11  |

**NB: *Always use vectorized expressions when dealing with vectors rather than writing loops for vectors as its much faster and less prone to errors***

## If else

-   Control structures are not vectorized, for instance if statements

-   Thus One can't write vectorized expressions for them

-   Check chunk below

    ```{r,error=T}
    x<-1:10
    if(x%%2==0) 5 else 15

    ```

-   To counter the issue we use the `ifelse()` function

```{r}
x<1:10
ifelse(x%%2==0,5,15)
```

## Vectorizing Functions

-   User defined functions have to be created in a way they work on vectors

-   That is, they take vectors as inputs and return vectors as output

-   This enables them to be used in vectorized expressions just like inbuilt functions

-   Consider the function below:

```{r}
f<-function(x,y){
  sqrt(x^y)
}
f(1:6,2)
f(1:6,2)
```

-   The `Vectorize()` is used when you want to coerce the use of vectorized expressions

-   Consider function that maps names in a class to their roles

```{r}
role_table<-list(Thomas="Instructor",
                 Henrick="Student",
                 Kristian="Student",
                 Randi="Student",
                 Heidi="Student",
                 Manfred="Student")
map_to_role<-function(name) role_table[[name]]

```

-   The function works when we call only one name

```{r}
map_to_role("Thomas")
map_to_role("Manfred")
```

-   It fails when we call the function with a vector/ when we call multiple names

```{r,error=T}
x<-c("Thomas","Henrick","Randi")
map_to_role(x)
```

-   Thus the function does not work on vectors

-   To make the function work on vectors, we use the `Vectorize()`function

```{r}
map_to_role<-Vectorize(map_to_role)
map_to_role(x)
```

-   The function initially failed due to use of `[[]]`

-   As previously stated `[[]]`are used to access the elements of the list directly

-   `[]`are used to access the list that contains elements we are interested in thus making it possible to ask for multiple names

```{r}
role_table <- list("Thomas" = "Instructor",
 "Henrik" = "Student",
 "Kristian" = "Student",
 "Randi" = "Student",
 "Heidi" = "Student",
 "Manfred" = "Student")
map_to_role<- function(name) role_table[name]
x<-c("Thomas","Henrik","Randi")
map_to_role(x)
```

-   But the results will come as a list and not as a vector of named values.

-   To offset it, we use the `unlist()`function that coerces lists to vectors

```{r}
map_to_role_2 <- function(names) unlist(role_table[names])
x <- c("Thomas", "Henrik", "Randi")
map_to_role_2(x)
```

**NB :** *When its not possible to write functions such that they take in vectors and output vectors we use `Vectorize()`*

-   The `[[]]`is a shortcut for accessing elements in a list.

-   It can also be used for recursive indexing

```{r}
x<-list(first=list(second="foo"),third="bar")
x[[c("first","second")]]
```

-   This can be written as:

```{r}
x[["first"]][["second"]]
x$first$second
```

# The apply Family

-   Vectorizing functions makes it easy to implicitly(indirectly) work with vectors, taking a vector as input and returning it as output

-   Not all functions are considered vectorized functions.

-   some (Ie; `sum()`,`mean()`) take sequence of vectors and return a single value

-   It requires explicit handling of sequence input in order to write such functions

-   R provides various ways to operate on sequences and call functions on them

-   Common functions used in R for such purposes include `apply` in their names

-   The apply functions are used to operate across data of data structures efficiently

## `apply`

-   Best explained using a matrix

-   `apply()` let's you specify if operation should be done on a row or column

-   It takes at leas t3 arguments:

    1.  Data from a data structure
    2.  Dimensions; 1 specifies row, 2 column and c(1,2) both
    3.  Function to be applied on data

```{r}
m <- matrix(1:9, nrow = 3, byrow = TRUE)
print(m)
#Apply sum function to each row
row_sums <- apply(m, 1, sum)
print(row_sums)
# Apply sum function to each column
col_sums <- apply(m, 2, sum)
print(col_sums)
```

-   **Apply paste function for each row**

```{r}
m<-matrix(1:6,nrow=2,byrow=T);m
apply(m,1,function(x) paste(x,collapse = ":"))
```

```         
The collapse argument is used in functions for concatenating such as paste()

It allows one to specify the delimiter that separates concatenated values

We apply the paste function to our rows thus values of each row are concatenated as individual expressions and the collapse argument specifies ":" as the separator of our concatenated values
```

```{r,warning=F,echo=F}
library("dplyr")
```

-   **Apply paste function for each column**

```{r}
m %>% apply(2,function(x) paste(x,collapse = ":"))

```

-   **Apply paste function for each row and column**

```{r}
m %>% apply(c(1,2),function(x) paste(x,collapse=":"))
```

-   The output of the chunks is 2-dimensional as we are marginalizing over two dimensions(rows and columns)

-   Consider a function that takes a vector and duplicates it by concatenating it with itself

-   If we apply it to our row or column we get an output for each row/ column that is a 2 dimensional

```{r}
m %>% apply(1,function(x) c(x,x))
```

-   When our function is applied to the rows of our matrix the result is a 2 column matrix

-   The 1^st^ columns contains the 1st row of the original column and its duplicate

-   The 2^nd^ column contains the 2nd row of the original column and its duplicate

-   Reason for this arrangement is because of the default column wise data input criteria of R

```{r}
m %>% apply(2,function(x) c(x,x))
```

-   When the function is applied to the matrix it results to a matrix with 3 columns

-   The 1^st^ column has the 1st column of the original matrix and its duplicate

-   The 2^nd^ column has the 2nd column of original matrix and its duplicate

-   The 3^rd^ column has the 3rd column of original matrix and its duplicate

-   Reason is same as above's

```{r}
m %>% apply(c(1,2),function(x)c(x,x))
```

## Rudia explanation

```{r}
x<-m %>% apply(c(1,2),function(x)c(x,x))
k<-dim(x)[3]
n<-dim(x)[2]
for(i in 1:n){
  for(j in 1:k){
    print(x[,i,j])
  }
}
```

-   In the event the user defined function takes additional arguments that are not in the data, you can define those arguments in the `apply()` function

```{r,error=T}
sumpow<-function(x,n) sum(x)**n
apply(m,1,sumpow)

```

-   The function `sumpow()`is not evaluated as the argument `n` has no value

-   This is offset by passing the value of `n` as an argument in the `apply()`function

```{r}
sumpow<-function(x,n) sum(x)**n
apply(m,1,sumpow,2)
```

-   For readability, the chunk can be written as

```{r}
apply(m,1,sumpow,n=2)
```

## `lapply`

-   It's used to map a function to a list

-   It takes a list as input and applies a specified function to each element of each list in the list and outputs a list of the same length

```{r}
(l<-list(1,2,3))
lapply(l,function(x) x^2)
```

-   If elements of a list are named the names are preserved for the output vector

```{r}
(l<-list(a=1,b=2,c=3))
lapply(l,function(x)x^2)
```

-   If you it a vector as input, it returns a list as output

```{r}
lapply(1:3,function(x)x^2)
```

-   If elements of the list are more complex than a single number, the function is still applied to the element of each list

```{r}
k<-list(a=1:3,b=4:6)
lapply(k,function(x)x^2)
```

-   Can also be written as:

```{r}
lapply(list(a=1:3,b=4:6),function(x)x^2)
```

## `sapply` and `vapply`

-   `sapply()` is used to map functions to each element of a vector

-   If given a list as input, it try to guess the output to return,(often a vector)

-   It will return a list when it fails to figure out the output you want

```{r}
sapply(1:3, function(x)x^2)
sapply(k, function(x)x**2)
```

-   `vapply()` operates as `sapply()` but has additional parameters to specify the output you want thus no guessing is done

-   The parameter is a value of the type of output you want

```{r}
vapply(1:3,function(x)x^2,3)
```

# Advanced Functions

## Special names

-   Special names on functions are used when, the name of the variable holding the user defined function is not accepted by R as a variable.

-   For instance you can't use the name of an inbuilt function as the name of a user defined function

-   One can also not use infix operators (used to evaluate values Ie; +,-,\< as functions names

-   To offset this, back-ticks are used

```{r,error=T}
if <-function(x)x^2
if(2)
```

```{r}
`if`<-function(x)x^2
`if`(2)
```

-   Using infix operators as normal function calls

```{r}
2+2 #infix function
`+`(2,2) #prefix operator

4:6*3-1 #infix operator
`-`(`*`(4:6,3),1) #prefix operator
```

-   Using function as infix operator

```{r,error=T}
`%or die%`<-function(test,msg) {
  if (!test) {
    stop(msg)
  }
}
x<-5
(x!=0)%or die%("x should not be zero")
x<-0
(x!=0) %or die% "x should not be zero"
```

## Infix operators

-   In R any variable that starts and ends with `% is considered and infix operator`

-   Inbuilt infix operators without this syntax include `+`,`*`,`/`,`-` e.t.c

-   Inbuilt operator with this syntax is the `%>%` of the `dplyr` package the pipe operator

```{r}
`%per%`<-function(prop,total) val= (prop/total)*100
(10%per%20) #infix operator function
(`%per%`(10,20)) #prefix operator

```

-   The expression on the left is the 1st argument of our infix operator function

```{r,error=T}
`%div%`<-function(a,b){
  if(a%%b==0) {return(TRUE)
}else return(FALSE)
}
20%div%10
`%div%`(20,10   )
```

## Replacement Functions

-   They are functions that 'pretend to modify variables'

-   For instance the `names()` used to name values of variables

```{r}
v<-1:4
names(v)<-c('a','b','c','d');v
```

-   R realizes you are assigning to a function call and looks for a function named `names<-`

-   It calls the function with the vector `v` and the vector of names

-   The result of the function call gets assigned to the variable `v`

-   This can also be expressed as

```{r}
v<-`names<-`(v,c('a','b','c','d'));v
```

-   Replacement functions are generally used to modify various attributes of an object

-   A function is referred to as a replacement function if its name has the suffix `<-`

-   A user defined replacement functions that performs the same as the `names()` function

```{r}
`foo<-`<-function(x,value){
  x$foo<-value
  x
}
`bar<-`<-function(x,value){
  x$bar<-value
  x
}
x<-c(foo=1,bar=2);x # a vector object
x<-list(foo=1,bar=2);x # a list object
x$foo #access the element in the list 'foo'
foo(x)<-3
x$foo
x$bar
bar(x)<-3
```

**Explanation of chunk above:**

-   We declare a replacement function `foo()` and `bar()`

-   Inside body of functions we use string/character indexing to declare a value of x for the variables `foo` and `bar`

-   The 2 functions each have 2 arguments:

    -   x, which is the object to create

    -   value, which is the element of the created object

-   We've defined our object x as a list

-   Our replacement function `foo(x)` is used to change the value in the list `foo` from 1 to 3

-   Same applies for our function `bar(x)`

-   This 2 replacement functions can also be called as:

```{r}
x<-`foo<-`<-3
x<-`bar<-`<-3

```

**RULES WHEN DECLARING REPLACEMENT FUNCTIONS**

1.  The parameter `value` should by default be named value. Usage of another variable names brings an error

```{r,error=T}
`foo<-`<-function(x,val){
  x$foo<-val
  x
}
x<-list(foo=1,bar=2)
foo(x)<-3
```

2.  You can give the object variable any name of your liking

```{r}
`foo<-`<-function(y,value){
  y$foo<-value
  y
}
x<-list(foo=1,bar=2)
x$foo
```

3.  The `value` parameter should always be last if there are more parameters in the function

```{r}
`modify<-`<-function(x,var,value){
  x[var]<-value
  x
}
x<-list(foo=1,bar=2)
modify(x,"foo")<-3
modify(x,"bar")<-4
x
```

# Functional Programming

## Anonymous functions

-   They are functions that are declared but not assigned any variable

-   Instead of having a function as:

```{r}
square<-function(x)x^2
```

-   You can have it as :

```{r}
function(x)x^2
```

-   Anonymous functions are used when you want:

1.  To use a one-off function that you don't need to give a name
2.  To create closure

-   Both cases are used when a function is used as an argument of another function or when returned from another function

**Case 1**

Compute sum of squares over row of a matrix using `apply()` function

```{r}
# without anonymous function
m<-matrix(1:6,nrow=3);m
sum_of_squares<-function(x)sum(x^2)
apply(m,1,sum_of_squares)
```

```{r}
# with anonymous function
apply(m,1,function(x)sum(x^2))
```

-   We can manipulate the power operator to square rows before finding the sum

```{r}
apply(m^2,1,sum)
```

**Case 2**

-   Using anonymous functions to create closures to write a function that returns a function

```{r}
#Without anonymous function
f<-function(x){
  g<-function(y) x+y
  g
}
```

```{r}
#with anonymous function
f<-function(x) function(y) x+y
```

## Functions Taking Functions as Arguments

-   The apply family best describes functions that takes other functions as its argument

-   Let's create a function that works like the `s/vapply()` but only takes inputs that satisfy a certain predicate/condition

```{r}
apply_if<-function(x,p,f){
  result<-vector(length=length(x))
  n<-0
  for(i in seq_along(x)){
    if(p(x[i])){
      n<-n+1
      result[n]<-f(x[i])
    }
  }
  head(result,n)
}
apply_if(1:8,function(x) x%%2==0,function(x)x^2)
```

## Function Returning Functions ( and Closures)

-   Closures enables functions remember the enviroment they were created in

-   Thus when a function is called, its able to access its variables and values even if its creation enviroment DNE

-   Closures are also created when you create a function inside another function and return it

-   The inner function can refer to parameters and local variables inside the outer function even after you have returned from it

-   Inner functions are used to specialize generic functions

-   Below is a generic power function specialized for squaring or cubing numbers

```{r}
power<-function(n) function(x) x^n
square<-power(2)
cube<-power(3)
x<-1:4
square(x)
cube(x)
```

## Filter, Map and Reduce

Filter, Map and reduce are patterns often used in functional programming

1.  `Filter()` is used to select cells that specify a given condition/predicate

```{r}
is_even<-function(x) x%%2==0
is_even %>% Filter(1:8)
Filter(is_even,1:8)
is_even %>% Filter(as.list(1:8))
Filter(is_even,as.list(1:8))

```

2.  **`map()`** functions work like `lapply()`

-   It applies a function to every cell/element of an object and returns a list of the result

-   Use `unlist()` to have result as a vector

```{r}
square<-function(x)x^2
Map(square,1:4)
```
