---
title: "CHAPTER1"
author: "ME"
date: "2024-04-26"
output: html_document
---

# **VECTORS**

A vector is a one dimensional data structure that stores data of the same type.\
An example includes:

```{r}
v<-1:5;v
```

# Accessing values of a vector

We can access either 1 individual element, a sequence of elements, a specified order of elements.

-   There are various ways of accessing elements of a vector. Such include using:

    1.  Integer vector as index.
    2.  Logical vector as index.
    3.  Character vector as index.

# 1.Integer vectors as index

## Accessing only 1 element.

A single element can be accessed by specifying its index inside square brackets as shown in the examples below.

\*Accessing:

-   Element in 1^st^ index:

```{r}
v[1]
```

-   Element in 2^nd^ index:

```{r}
v[2]
```

-   Element in 3^rd^ index:

```{r}
v[3]
```

## Accessing a sequence of elements:

A sequence of elements can be accessed using the colon and interval of integers to be accessed,\
Ie - [a:b] ;a is the lower and b the upper limit

\*Accessing:

-   All elements:

```{r}
v[1:5]
```

-   1^st^ four elements:

```{r}
v[1:4]
```

-   Last 3 elements:

```{r}
v[3:5]
```

## Accessing specified number elements of intrest:

We can access a specified number of elements using the c() Ie:

\*Accessing:

-   1^st^, 3^rd^ and 5^th^ elements:

```{r}
v[c(-1,-3,-5)]
```

## Accessing all elements other than those specified

Just like accessing elements in the previous scenarios, we can access all elements other than that specified by using the negative sign. Ie;

\*Accessing:

-   All elements other than the 3^rd^ one:

```{r}
v[-3]
```

-   All elements other than the 1^st^, 3^rd^ and 5^th^:

```{r}
v[c(-1,-3,-5)]
```

**NB:** Mixing of negative and positive index is not allowed

## 2.Logical vector as index

Element who's position is deemed true are accessed

\*Accessing:

-   1^st^ element

```{r}
v[c(T,F,F,F,F)]
```

One can access other elements using the same idea

## 3.Character vector as index

It only works if vectors are named.

There are two ways of naming vectors: a. Assigning names to vectors manually, Ie;

```{r}
x<-c('first'=3,'second'=5,'third'=10);x
x['first']
x[1]
x[c('first','third')]
```

b.  Using the names() function, Ie;

```{r}
names(v)<-c('d','e','f','g','h');v
v['e']
v[c('d','f','h')]
v[c(1,3,5)]
```

## Recycling process of vectors

The recycling process is performed in events where we're dealing with a short and long vectors.\
When performing executions with such vectors, R repeats the values of the shorter vector till they equal that of the longer vector

```{r}
(x<-1:3)
(y<-4:9)
x+y
y+x
```

## FUNCTIONS

They are of two types

1.  In-built functions
2.  User-defined functions

In-built functions are those that are already defined by R.\
They are readily available for execution

## User-defined functions

They are functions that are created by the user to perform some specific tasks.\
We'll have a number of examples of such functions and what they do.

*1. function that gives square of a number or a sequence of numbers*

```{r}
square<-function(x) x^2
square(2)
square(1:5)
```

*2. Function that get squares and subtracts from a single vector*

```{r}
square_subtract<-function(x,y) x^2-y
square_subtract(2,2)
square_subtract(1:5,rev(1:5))
```

**NB:** If body of function has an inline expression, there's no need of curly brackets like in the above examples.

# CONTROL STRUCTURES

-   We have various types of control structures:
    1.  If statements\
    2.  If else statements\
    3.  Ifelse() function\
    4.  If...else if statements\
    5.  Nested if...else statements\
    6.  Loops:
        a.  For loops\
        b.  While loops\
    7.  Return statements:
        -   break\
        -   next

## 1.if statements

Tests one condition as true or false.\
If TRUE the expressions inside curly brackets are executed.\
If FALSE the program stops.\
Examples:

The program below stops as condition is False

```{r}
if(2>3) 'false'
```

The program below has it's expression executed as condition is True

```{r}
if(3>2)'true'
```

Since condition is False, program stops and the value of x is unchanged

```{r}
x<-"foo"
if(2>3) (x<-"bar") #condition is false
```

Since condition is True, x is assigned a new value "baz"

```{r}
if(3>2) (x<-"baz") #condition is true
```

**NB:** If there's inline expression like those above, there's no need of curly brackets

## 2. if else statements

Tests one condition as true or false if TRUE, expressions in the if statement are executed.\
If false, expressions in the else statement are executed.\
Examples:

```{r}
if(2>3) "bar" else
"baz"
```

# 3. $ifelse()$ function

It is used when dealing with vectorized expressions( A vector with at least 1 value) The if, if..else, if...else if and the nested if... else if can only deal with vectorized expressions when loops are introduced. To avoid this, the ifelse() function is used Example:

```{r}
x<-1:5
ifelse(x>3,"bar","baz" )

```

## FUNCTIONS and IF STATMENTS

We're going to look at some examples.\
Consider the function:

**Example 1**

```{r}
m_s<-function(x){
  if(x%%2==0) x^2 else x
}
m_s((4))
m_s(3) #condition is false, odd number displayed
```

*The above function checks if a number is even.*\
*If so, it squares that number and prints it.*\
*Otherwise it prints that number as it is.*\
*However the function is not eligible for a sequence of numbers.*\
**NB:** To have the code accommodate vectorized expressions we coerce it into a vector, Ie;

```{r}
maybe_square<-Vectorize(m_s)
maybe_square((1:5))
```

**Example**

```{r}
m_s<-function(x){
  ifelse(x%%2==0,x^2,x)
}
m_s(4)
m_s(3)
m_s(1:5)
```

*The above code checks if number is even or odd.*\
*If even, it squares the number and prints it.*\
*Otherwise, it prints number as it is*\
*Function caters for vectorized expressions*

## 6. LOOPS

Loops are used to iterate a sequence of operations or commands without need to write code for each command

## a. For loops

-   Perform a finite number of iterations on given commands.
    -   Example 1:

        ```{r}
        x <- 6:10
        total <- 0
        for (i in x) total = total + i
        total
        ```

        -   The loop iterates over the values of x.\
        -   In each iteration, it assigns the value of x to i.\
        -   An expression inside the loop is used for summation purposes.\
        -   In each iteration, it adds the variable i to a variable total whose initial value is 0.\
        -   The summation result is stored in total, which is updated in each iteration.\
        -   We print total outside the loop, displaying only the final sum.

    -   Example 2:

        ```{r}
        x<-6:10
        t=0
        for(i in seq_along(x)){
          y<-x[i]
          t=t+y
        };t
        ```

        -   The seq_along() function is used to run throught the indices of a vector \*the variable y is assigned values of x that are accessed through indexing
        -   The loop is same as that above other than those 2 deviations

    -   Example 3:

        ```{r}
        # Your code for Example 3 goes here
        ```
