---
title: "CHAPTER4-DATA VISUALIZATION"
date: "2024-05-02"
output:
  pdf_document: default
  html_document: default
---

Data visualization is the use of plots and graphs to study data\
Let's begin with the basic plot

## Basics Graphics

The `graphics` package has some functions used in basic plotting\
There's no need to load the package `library("graphics")` as it's loaded when R starts.\
TO get most of the functions used run the chunk below

```{r,eval=FALSE}
library(help="graphics")
```

The code opens a documentation that gives a list of functions used in basic plotting\
1. **Scatter plot and line plot** The simplest plot is the scatter plot\
- A scatter plot is a 2 dimensional plot that shows the relationship of 2 continuous variables where each data point is represented by a single dot

```{r}
    x<-rnorm(50)
    y<-rnorm(50)
    plot(x,y)
```

Let's say we want to have a plot from a `dataframe`\
R is not able to know which column it's supposed to plot on right and left axis\
There are 2 ways to offset this,

```{r,echo=F,warning=F}
library("magrittr")
```

1.  **Accessing the columns manually**

```{r,warning=F}
    data(cars)
    cars %>% plot(cars$speed,cars$dist,data=.,
                  main="Cars data",
                  xlab="Speed",ylab="Stopping distance")
```

2.  **Using %\$% `exposition` pipe to access columns**

```{r}
     cars %$% plot(speed,dist,
                  main="Cars data",
                  xlab="Speed",ylab="Stopping distance") 
```

```         
-   `main` is used to name our plot  
-   `xlab` gives name(labels) the x-axis  
-   `ylab` labels the y-axis  
-   By default plot shows data as points, this can be changed using arguments  
-   For instance we can have our data shown as lines by adding the argument `type=h`  
```

```{r}
    cars %$% plot(speed,dist,main="Cars data",type="h",
                  xlab="Speed",ylab="Stopping distance") 
```

2.  **Histogram**\
    Its a 2 dimensional plot that displays the frequency distribution of data\
    The x-axis is for the data while the y-axis is for the frequency\
    The `hist()` function is used to plot one

```{r}
cars%$%hist(speed,breaks=10)
```

*NB:* *A `plot()` is a generic function. This implies it has different functionality based on the arguments you give it*

3.  **A simple linear regression curve**\

-   The `lm()` function is used to fit linear models
-   It has the argument `y~x`which shows the relationship between x and y where y is the response(dependent) variable and x the predictor(independent) variable

```{r}
cars %>% lm(dist~speed,data=.) %>% plot
```

The chunk will give several summary plots for visualizing quality of the linear fit

**More arguments in `plot` include `lines()points()` that add lines and points respectively to existing plot**

```{r}
data(longley)
str(longley)
longley %>% plot(Unemployed~Year,data=.,type="l")
longley %>% lines(Armed.Forces~Year,data=.,col="blue")
```

-   `?longley` gives description of the data set
-   We've plotted the graphs of unemployment against year and armed forces against year on the same plot by using the `lines()` function
-   However the all values of `Armed.Forces` are not fitted in our graph
-   This is because `plot()` takes \`Unemployment as the y-axis
-   To offset it we use the argument `ylim()` to fit the 2 variables as the y-axis

```{r}
longley %>% plot(Unemployed~Year,data=.,type="l",
                 ylim=range(c(Unemployed,Armed.Forces)))
longley %>% lines(Armed.Forces~Year,data=.,col="blue")
```

**Fitting line of best fit**

The `abline()` function is used to plot straight lines Together with `lm()` the can be used to plot a line of best fit

```{r}
cars %>% plot(dist~speed,data=.)
cars %>% lm(dist~speed,data=.) %>% abline(col="red")
```

**plotting `iris` data with data points of different shapes according to species**

```{r}
shape_map<-c("setosa"=1,
             "versicolor"=2,
             "virginicia"=3)
iris %$% plot(Petal.Length~Petal.Width,
              pch=shape_map[Species])
```

-   `pch` is used to add points in a plot

-   points are added as shapes which are defined using integers

-   use `?pch` to check out more about integers and their assigned shapes

-   The functions of `graphics` package are used for low level plotting

-   We will graduate to the `ggplot2` package that is used for high level plotting

## The Grammar of Graphics and `ggplot2` package

[**CLICK HERE**](https://ggplot2-book.org/getting-started) to access documentation of the package. Use `library(help="ggplot")` to get a list of functions of this package Let's install and load the package

```{r,eval=F}
install.packages("ggplot2")
```

```{r,warning=F,comment=NA}
library("ggplot2")
```

The package gives a domain specific language for plotting just like `dplyr` that gives a domain specific language for manipulating data

# Using `qplot()`

Works like `plot()` but is better\
By default, it plots scatter plots

```{r,warning=FALSE}
cars %>% qplot(speed,dist,data=.)
```

**1. Scatter plot**\
**NB:** `qplot()` creates a `ggplot` object before plotting unlike the `plot()` which plots directly Thus when you print this object the output is the graph to be plotted. ie,

```{r,warning=FALSE}
p<-cars %>% qplot(speed,dist,data=.)
p
```

This is equivalent to:

```{r,warning=FALSE}
p<-cars %>% qplot(speed,dist,data=.)
print(p)
```

With `qplot()` it simpler to plot iris data with data points of different colors based on species than it was with `plot`

```{r}
iris %>% qplot(Petal.Width,Petal.Length,
               color=Species,data=.)
```

## 2. Histogram

By giving one argument to the `qplot()` it plots a histogram

```{r}
cars %>% qplot(speed,data=.,bins=10)
```

The `bins` argument is used to specify intervals.\
If not specified, R will by default choose a suitable interval

In the event you wanted a density plot, you can ask R for it\
A **Density** plot displays frequency distribution of data using curves

```{r}
cars %>% qplot(speed,data=.,geom="density")
```

**`geom`** stands for geometry and is used to determine how data should be visualized or displayed\
One can add more that one geometry to a plot ie;

-   lines
-   bar plot
-   violin plots
-   etc

## Using Geometries

```{r}
cars %>% qplot(speed,dist,data=.)
```

To create the plot above using geometries

-   Create a ggplot object
-   Map your variables from the data frame,
    -   speed mapped to x-axis
    -   dist mapped to y-axis
-   Plot the data as points

```{r}
ggplot(cars) + geom_point(aes(x=speed,y=dist))
```

In the chunk above:

-   We've created an object using `ggplot()` and given it it's input as `cars` data set
-   We specified our x and y values be plotted as points using `geom_point` argument
-   Using `aes` that stands for aesthetics, we map speed to x-axis and dist to y- axis for the `geom_point` part
    -   Aesthetics are used to map from data to graphics
-   In the event we want `aes` to be shared with different functions, we set it in the `ggplot()`
    -   By doing so we don't have to specify it for all functions we want with the same `aes`

```{r}
ggplot(cars,aes(x=speed,y=dist)) + geom_point()
```

-   

    -   is used to string together a series of `cmds` used to modify the `ggplot()`object
    -   In our codes above it's used to combine `ggplot()` and `geom_point()`
    -   `+` is the `%>%` version of `ggplot2` package

-   since `ggplot()` takes a data frame as it's 1^st^ argument, we use `%>%` when plotting

```{r}
cars %>% ggplot(aes(x=speed,y=dist)) + geom_point()
```

For the `iris` data set the code chunk below is used to plot it's scatter plot

```{r}
iris %>% qplot(Petal.Width,Petal.Length,
              color=Species,data=.)
```

The equivalent using geometries is

```{r}
iris %>% ggplot +
  geom_point(aes(x=Petal.Width,y=Petal.Length,
             color=Species))
```

We can have the aesthetics inside the `ggplot()`

```{r}
iris %>% ggplot(aes(x=Petal.Width,y=Petal.Length,color=Species)) +
  geom_point()
```

**NB:** in the chunks above, we have the argument `color` inside `aes` as it is dependent on another variable.\
In the event we want a specific color, we'll have the argument outside `aes`

```{r}
iris %>% ggplot +
  geom_point(aes(x=Petal.Width,y=Petal.Length),
             color="red")
```

```{r}
iris %>% ggplot(aes(x=Petal.Width,y=Petal.Length)) +
  geom_point(color="red")
```

**`qplots()` for histogram and density plots**

```{r}
cars %>% qplot(speed,data=.,bins=10)
cars %>% qplot(speed,data=.,geom="density")
```

**`ggplots()` for histogram and density plots**

```{r}
cars %>% ggplot(aes(x=speed)) +
 geom_histogram(bins=10)
```

```{r}
cars %>% ggplot(aes(x=speed)) +
   geom_density()
```

**combined plot for histogram and density using `ggplot()`**

```{r,warning=F}
cars %>% ggplot(aes(x=speed,y=..count..)) +
  geom_histogram(bins=10) +
  geom_density()
```

By setting `y=..count..` you tell both geometries to use counts as the y-axis\
To get densities instead you use `y=..density..`

```{r}
cars %>% ggplot(aes(x=speed,y=..density..)) +
  geom_histogram(bins=10) +
  geom_density()

```

**Fitting a line of best fit**

```{r}
cars %>% ggplot(aes(x=speed,y=dist)) +   geom_point() + geom_smooth(method = "lm")

```

-   The `geom_smooth()` is used to plot a smooth curve(linear curve)

-   You define the type of smooth curve in the `method` argument

-   Other common methods include:

```         
-    "loess" for local regression (which is the default method),(plots a non linear curve)

-    "gam" for generalized additive models, and

-    "glm" for generalized linear models
```

### Using at-least one geometry in a single plot

-   Just like in the chunks above, you can have more than one geometry in the plot as long as they are separated by `+`

-   You can also specify if you want different `aes` for each geometry function or a similar for each function as shown above

-   Some examples include:

```{r}
longley %>% ggplot(aes(x=Year)) +   geom_line(aes(y=Unemployed)) +   geom_line(aes(y=Armed.Forces),color="blue")

```

-   Since our variables share the x-axis/x-value we define `aes(x=Year)` in the `ggplot()` object

-   We define different y-values in their respective geometry functions as they are not same

-   **NB:** The y-axis is adjusted to fit both lines thus no need of `ylim=range()`- the y-limit function

```{r}
longley %>% ggplot(aes(x=Year)) +   geom_point(aes(y=Unemployed)) +   geom_point(aes(y=Armed.Forces),col="blue") +   geom_line(aes(y=Unemployed)) +   geom_line(aes(y=Armed.Forces),col="blue")

```

Plotting 2 variables with different aesthetics is a trick that mostly works but isn't that efficient

In the chunk above we are representing our variables `Unemployed` and `Armed.Forces as 2 different measures`

Thus it's impossible to split the plot as 2 sub-plots when we need

To be able to represent the variables as 1 measure and thus able to split them into subsets , we'll employ the `gather()`

The $gather()$ makes wide data to be long data.

It has the arguments $key$ that stores the variables of interest and $value$ which stores the value of the variables

```{r}
library(tidyr)
```

```{r}
longley %>% 
  gather(key,value,Unemployed,Armed.Forces) %>% 
  ggplot(aes(x=Year,y=value,color=key)) +
  geom_line() +
  geom_point()
```

It is now possible to split our plot into 2 sub plots using `facet`

```{r}
longley %>% 
  gather(key,value,Unemployed,Armed.Forces) %>% 
  ggplot(aes(x=Year,y=value,col=key)) +
  geom_line() +
  facet_grid(key~.)
```

```{r}
longley %>% gather(Key,Attribute,Unemployed,Armed.Forces) %>% 
  ggplot(aes(x=Year,y=Attribute,col=Key))+
  geom_line()+
  geom_point()+
  facet_grid(Key~.)
```

## Facets

They are used to split plots into subplots

Facets can be specified using the functions `facet_grid()` and `facet_wrap()`

`facet_grid()`

:   -   creates subplots based on two categorical variables

    -   **syntax:** rows\~columns

    -   By default `ggplot` puts all values in the same axis when creating facets using `facet_grid`

    -   facets within a column will always have same x-axis whilst those within a row will have same y-axis

    -   Sharing same axis by different variables isn't that effective as they have they have different ranges. R will not consider this.

    -   To offset this, we use the `scale` argument to give each variable its range

    <div>

    ```{r}
    iris %>% 
      gather(Measurement,value,-Species) %>% 
      ggplot(aes(x=Species,y=value)) +
      geom_boxplot() +
      facet_grid(Measurement~.)
    ```

    </div>

    `scale` gives each variable its range

    `scale="free_y"` is used to specify that scale of the y- axis in each facet

    <div>

    ```{r}
    iris %>% gather(Measurement,value,!Species) %>% 
      ggplot(aes(x=Species,y=value,col=Measurement)) +
      geom_boxplot() +
      facet_grid(Measurement~.,scale="free_x")

    ```

    </div>

    -   By default all `facets` have same size. You can modify this using the space variable.

    -   Its useful for categorical values if one facet has many levels than the others

    -   Labels used are taken from factors in the variables used to construct the facet

    -   In the event you want to modify labels, the `labeller` argument of `facet_grid()` is used

    -   The `labeller` takes a function (ie; `labeller()`) as an argument responsible for making labels

    -   Let's modify the iris label above to have white space as delimiter instead of periods

    <div>

    ```{r}
    label_map<-c(Petal.Width="Petal width",
                 Petal.Length="Petal Length",
                 Sepal.Width="Sepal Width",
                 Sepal.Length="Sepal Length")
    iris %>% gather(Measurement,value,!Species) %>% 
      ggplot(aes(x=Species,y=value,col=Measurement)) +
      geom_boxplot() +
      facet_grid(Measurement~.,
                 scale="free_y",
                 labeller = labeller(Measurement=label_map ))

    ```

    </div>

`facet_wrap()`

:   -   creates subplots based on one categorical variables

    -   **syntax** : \~variables

# Scaling {#sec-scaling}

Scaling refers to the process of adjusting the range of values of a data set to make them more suitable for analysis

scales tell `ggplot` how dimensions should be visualized

1.  **Naming the x and y axis**

```{r}
cars %>% 
  ggplot(aes(x=speed,y=dist)) +
  geom_point() +
  geom_smooth(method='lm') +
  scale_x_continuous("Speed") +
  scale_y_continuous("Stopping distance")
```

`scale_x/y_continuous` are used to control axis graphics for continuous data

2.  **Setting break points**

-   We can set break point to be shown if we wanted tick marks for each year of the `longley` data set

```{r}
longley %>% gather(key,value,Unemployed,Armed.Forces) %>% 
  ggplot(aes(x=Year,y=value)) +
  geom_line() +
  scale_x_continuous(breaks =1947:1962) +
  facet_grid(key~.)
```

3.  **Reversing**

To reverse and axis we use the `scale_x\y_reverse()` function

For instance in the `cars` data set you can show speed of cars is decreasing instead of increasing

```{r}
cars %>% ggplot(aes(x=speed,y=dist)) +
  geom_point() + geom_smooth(method="lm") +
  scale_x_reverse("speed") +
  scale_y_continuous("Stopping distance")
```

4.  **Naming labels**

In the event you're dealing with discrete data, `scale_x/y_discrete()` is used to define dimensions of our plots

**NB:** A factor is considered to be discrete data, so when used in an axis we employ `scale_x/y_discrete` to define our dimension.

```{r}
iris %>% ggplot(aes(x=Species,y=Petal.Length)) +
  geom_boxplot() +
  geom_jitter(width = 0.1,height = 0.1)
```

`geom_jitter()`has been used to show individual data points in our box plot

In modification of the x-axis, we will use the `scale_x_discrete()`

```{r}
iris %>% 
  ggplot(aes(x=Species,y=Petal.Length))+
  geom_boxplot() +geom_jitter(width=0.1,height=0.1) +
  scale_x_discrete(labels=c("setosa"="Setosa",
                            "versicolor"="Versicolor",
                            "virginica"="V irginica"))
```

5.  **Coloring plots**

The argument `col` or `color` is used to color the edges of objects in a plot whilst the argument `fill` is used to color this objects

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,color=Species)) +
  geom_boxplot() +
  facet_grid(Measurement~.,scale="free_y") +
  scale_x_discrete(labels=c("setosa"="Setosa",
                            "versicolor"="Versicolor",
                            "virginica"="Virginicia"))
```

The code is used to color the outlines

The next code will be used to color the inside the object

```{r}
iris %>% 
  gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,fill=Species)) +
  geom_boxplot() +
  facet_grid(Measurement~.,
             scale = "free_y",
             labeller = labeller(Measurement=label_map
                                 ))
```

**Modifying colors**

The `scale_fill_manual()` function can be used to modify colors

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,fill=Species)) +
  geom_boxplot() +
  scale_fill_manual(values=c("red","green","blue")) +
  facet_grid(Measurement~.,space="free_y",
             labeller = labeller(Measurement=label_map))
```

Setting colors can be tricky especially if you don't know the right colors to set to suite color blind people

This can be offset using the `scale_fill_brewer()` that let's you work with color palettes.

A color palette is a combination of colors choose to work together

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,fill=Species)) +
  geom_boxplot() +
  scale_fill_brewer(palette="Greens") +
  facet_grid(Measurement~.,scale="free_y")
```

## Theme and other Graphics Transformation

Themes are used to define how the final plot will look like

To define how final plots look like, we have to modify our themes

The default theme is given by the function

```{r}
theme_set(theme_bw())
```

`theme_bw()` sets up the looks of figure

`coord_flip()` is used to flip(interchange) the x and y-axis rather than changing it from the aesthetics

1.  Code with original x and y axis

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,fill=Species)) +
  geom_boxplot()+
  scale_x_discrete(labels=c("setosa"="Setosa",
                            "versicolor"="Versicolor",
                            "virginicia"="Virginicia")) +
  scale_fill_brewer(palette = "Greens")+
  facet_grid(Measurement~.,switch = "y",
             labeller = labeller(Measurement=label_map))
```

2.  Code with x and y axis interchanged using aesthetics

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=value,y=Species,fill=Species)) +
  geom_boxplot()+
  scale_x_discrete(labels=c("setosa"="Setosa",
                            "versicolor"="Versicolor",
                            "virginica"="Virginica")) +
  scale_fill_brewer(palette="Greens") +
  facet_grid(Measurement~.,switch="y",
             labeller=labeller(Measurement=label_map))
```

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,fill=Species)) +
  geom_boxplot() +
  scale_x_discrete(labels=c("setosa"="Setosa",
                   "versicolor"="Versicolor",
                   "virginica"="Virginica")) +
  scale_fill_brewer(palette="Greens") +
  facet_grid(Measurement~.,switch="y",
             labeller=labeller(Measurement=label_map))+
  coord_flip()
```

While `scale="free_y"` is used to specify the scale of the y-axis of each facet, `switch=y` is used to specify that scales shouldn't be shared on the y-axis

When `scale="free_y'` is used together with `coord_flip="y"` only the 1^st^ facet will have it's scale and the rest will share a scale

We want to get rid of the background in the labels and also move the legend to the top

The functions `` theme(strip.background=element_blank() and `theme(legend.position="top ``

```{r}
iris %>% gather(Measurement,value,!Species) %>% 
  ggplot(aes(x=Species,y=value,fill=Species)) +
  geom_boxplot()+
  scale_x_discrete(labels=c("setosa"="Setosa",
                            "versicolor"="Versicolor",
                            "virginica"="Virginica")) +
  scale_fill_brewer(palette = "Greens") +
  facet_grid(Measurement~.,
             switch="y",
             labeller=labeller(Measurement=label_map))+
  coord_flip()+
  theme(strip.background=element_blank())+
  theme(legend.position="top")

```

### Figures with multiple plots

`ggplot` does not directly support combining multiple plots

In combination with the underlying graphics system `grid` thus can be achieved

however `grid` has many low-level-tools for modification of graphics

The `gridExtra` package can be used to combine plots for high-level-functions

To combine plots:

1.  **Make individual plots**

```{r}
petal<-iris %>% ggplot()+
  geom_point(aes(x=Petal.Width,y=Petal.Length,color=Species))+
  theme(legend.position="none")
sepal<-iris %>% ggplot(aes(x=Sepal.Width,y=Sepal.Length,color=Species))+
  geom_point()+
  theme(legend.position = "none")
petal
sepal
```

2.  **Install and Load the `gridExtra`** **packages**

```{r,eval=F,warning=F}
install.packages("gridExtra")
```

```{r,warning=F}
library("gridExtra")

```

3.  Us ethe `grid.arrange()`to create a grid of plots in the plots we've created

```{r}
grid.arrange(petal,sepal,ncol=2)
```

Another package that can be used to combine plots is the `cowplot` package

It redefines the default theme of the `ggplot` package

```{r,eval=F,warning=F}
install.packages("cowplot")
```

```{r,warning=F}
library("cowplot")


```

```{r}
plot_grid(petal,sepal,labels=c("A","B"))
```
