---
title:  CHAPTER_3   
        Data_Manipulation   
output: html_document   
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

Data manipulation is the process of sorting,filtering
aggregating,transforming and summarizing data sets to meet certain user
satisfaction

We'll have to install and load some important packages used in data
manipulation and they are included in the chunk code below:

```{r,eval=FALSE}
install.packages('magrittr')
```

```{r,warning=FALSE}
library('magrittr')
```

```{r,eval=FALSE}
install.packages('ggplot2')
```

```{r,warning=FALSE}
library('ggplot2')
```

More will be interacted with later

R has some inbuilt data sets that can be used in studying data
manipulation.\
These data sets come in their data set packages.

# Loading data in R

The **data()** function is used to load data sets in R

-   Example 1:

```{r,warning=F}
data(cars)
head(cars) 
str(cars)
cars%>%qplot(speed,dist,data=.)
```

-   We load the data sets cars in R
-   `head()` is used to output 1^st^ 6 rows
-   `str()` shows the structure of the data set cars
-   `qplot()` is used in plotting

The *`mlbench`* data set package contains data sets for machine learning
and we'll be using it

```{r,eval=F,warning=FALSE}
install.packages('mlbench')
```

```{r,warning=FALSE}
library('mlbench')
```

## Data reviewing in R

```{r,echo=F,warning=FALSE}
library('knitr')
```

```{r}
cars%>%head(3)
cars%>%tail(3)
cars%>%summary%>%kable(.)
```

-   `head()` displays 1^st^ n rows, in our case 1^st^ 3 rows.
-   `tail()` displays last n rows, in our case 3 last rows.
-   summary() gives the summary of given data set
    -   summary given is for each variable(col) with observations
-   `kable()` is a function from *`knitr`* package that draws tables

Consider the data set iris

```{r}
data('iris')
iris%>%summary()
iris%>%summary%>%kable
```

*NB:*

-   The summary you get depends on the type of columns you have.
-   Numerical data is summarized by their `quantiles` whilst `boolean`
    is summarized by counts of each category or True/False values

# Reading data in R

We have various functions for reading(importing) data to R.\
The code "`?read.table()`" has a list of such functions.\
The functions include:

1.  **read.table()** :
    -   Data are separated by a white space
    -   for reading files with tabular data(data stored in tables)
2.  **read.csv()** :
    -   CSV stands for comma separated values
    -   Data are separated by commas
    -   for reading files with tabular data
3.  **read.xlsx()** :
    -   for reading data stored in excel
4.  **read.delim()** :
    -   for reading files with TSV data (Tab separated values data)

    -   Delimited data is that which is separated by a specific
        delimiter.

    -   Eg: a csv data whose specific delimiter is a comma.

    -   The function is effective for reading TSV data **NB:** Data
        loaded in R using the above functions is stored as a data frame.

Some of the arguments used:(more arguments can be found in "
`?read.table()` "):

1.  **`header`** :

-   Function is defined as True or False
-   If file has columns, then argument is defined as True and vice-versa
-   **NB:**
    -   If set as True, R uses the 1^st^ row to set column names
    -   Otherwise the 1^st^ line is used as the 1^st^ row of the
        `dataframe`

2.  **`col.names`**:

-   If `header=F`, argument is used to name columns of data frame
-   A vector of string with a string for each column is defined

3.  **`dec`**:

-   argument specifies whether a "," or "." should be used for decimal
    numbers

4.  **`comment.Char`**:

-   By default "\#" is used for comments
-   If "\#" is used in data the argument is used to change it to avoid R
    assuming that data are comments
-   same applies if another symbol is used for comments in the data

5.  **`stringAsFactors`**:

-   By default, columns with strings are interpreted as factors
-   The argument is set to False when column strings are not factors

6.  **`colClasses`**:

-   Allows one to specify the type each col should have, ie;
    -   string or factor
-   specification can only be done for all cols and not 1

## Examples of Reading and Formatting Data sets

1.  **`Breast Cancer Dataset`**

-   Data is obtained from *`mlbench`* package

```{r,echo=F}
  library("magrittr")
  library('mlbench')
```

```{r}
    data(BreastCancer)
  BreastCancer%>%head(3)
```

-   The code above gives us a glimpse if this data set

-   We want to read the data from it's original website and store it in
    a variable, `data_url`

-   [Here's the link to the
    data](https://archive.ics.uci.edu/dataset/15/breast+cancer+wisconsin+original)

-   There are 2 ways in reading this data into R:

    A.Loading data indirectly from website by storing it in a local
    file.

    -   The con here is that when sharing other people can't access the
        file as it's in the directory of the machine that originally
        wrote the code.
    -   One can also make mistakes in the steps of downloading the file.

    B. Loading the data directly using the website

    -   The con here is that the website can be brought down and thus
        data is lost
    -   The administrators of the server can also modify data corrupting
        the original version we needed

**A. Loading data indirectly from website by storing it in a local
file**

-   The steps here include:

1.  Storing data set's website URL (link) to a variable `data_url`

```{r,eval=T}
  data_url<-"https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data"
```

2.  Use the `readLines()` to check delimiter used to separate values

```{r,eval=F}
  line_s<-readLines(data_url)
  line_s[1:5]
```

```         
- The `readLines()` function corerce our data into a vector of lines
    - We use it to check the separator of our data.
    - Our data is CSV thus `read.csv()` is used.
    - We store the vector in a variable called line_s
- We then access the 1^st^ 5 rows which are stored in line_s inorder to see our delimiter
```

3.  Use `writeLines()` to store a list of strings to a file.

```{r,eval=F}
writeLines(line_s,con="C:/Users/ganje/OneDrive/Documents/project/R_work/data/raw-breast-cancer.csv")
```

-   The `writeLines()` is used to write a list of strings to a file
-   Our list of strings is stored in the vector string `lines_s`
-   The argument `con` let's you specify the path where our file is
    stored

4.  Use `read.csv()` to read our data from the file it's stored in

```{r}
raw_breast_cancer<-read.csv("data/raw-breast-cancer.csv",col.names=names(BreastCancer))
raw_breast_cancer %>% head(3)
```

```         
- Compare `BreastCancer` and our imported data `raw-breast-cancer`
- Are they equal?
```

**B. Loading the data directly using the website**\
Steps:

1.  Use our variable `data_url` that has our stored link
2.  Use `read.csv()` to read data from our link

```{r,eval=F}
raw_breast_cancer<-read.csv(data_url,header=F,col.names = names(BreastCancer))
raw_breast_cancer %>% head(3)
BreastCancer %>% head(3)
```

-   Our imported data set `raw_breast_cancer` and the data set from
    *`mlbench`* `BreastCancer` are all similar apart from the column
    `class`
-   `raw_breast_cancer` has observations for `class` as 2 and 4 (2 for
    `bening`, 4 for `malignant`)
-   Our goal is to make the data sets same thus some modifications are
    done to the imported data set and they include:

1.  Translating numbers(2 and 4) into factors.
    -   We'll translate the numbers to strings then to factors
    -   We don't want to lose original data of imported data set thus we
        create a new variable.

```{r}
    formatted_breast_cancer<-raw_breast_cancer
```

```         
- Using `ifelse()` we map(transform) numbers to string
- we'll create a function to reduce variables we have
```

```{r}
    map_class<-function(x){
      ifelse(x==2,"bening","maligant")
    }
    mapped<-formatted_breast_cancer$Class %>% map_class
    mapped %>% table
```

-   `table()` is used to show the number of times a value in a vector
    appears

```{r}
    formatted_breast_cancer $Class %>% unique
```

-   `unique()` removes duplicates in vectors. Shows the unique elements
    in a vector

-   *NB:* We don't want to assume that the only values are 2 and 4 (ie,
    there are no unexpected values). Thus we'll have to set the
    unexpected values as NA

-   To do this:

    1.  Create a table `dict` to map through values.

-   Nested `ifelse` can be used but in our case we have many different
    possible values thus cumbersome to use them

-   We've named our table `dict` to avoid confusion with `table()`

-   The table dictionary `dict` will be used to look up words

-   `dict` should be a vector we can use strings(names) as indices

    2.  When we use a vector as a table, we need to have it's keys(for
        the table) as strings
    3.  Thus the numbers from the vector we want to map from should be
        converted to strings before looking them up in the dictionary.
        Below code implements this

    ```{r}
        dict<-c("2"="benign","4"="malignant")
        map_class<-function(x) dict[as.character(x)]
        
        mapped<-formatted_breast_cancer$Class %>% map_class
        mapped %>% table
    ```

    If we look at our vector `mapped`, we see that its a named vector
    and this is not our desired result

    ```{r}
        mapped[1:5]
    ```

-   We'll use the `unname()` function to offset this

```{r}
    mapped %<>% unname
    mapped[1:5]
```

-   The assignment pipe `%<>%` takes output from LHS to the RHS for
    execution. It then takes the output from RHS to LHS

-   The tee pipe `%t>%` takes output from LHS to RHS but does not store
    it

-   The pipe operator `%>%` takes output from LHS to RHS and stores it

-   The exposition pipe \` %\$%

    4.  The final step is to make the values of class factors

```{r}
    formatted_breast_cancer$Class %>% 
      map_class %>% 
      unname %>% 
      factor(levels=c("benign","malignant")) %>% 
      levels
```

**The entire reading of data and formatting is shown below:**

```{r,echo=F}
library("magrittr")
library("mlbench")
data("BreastCancer")
```

```{r,eval=FALSE}
# get data from website using tinyurl
data_url<-"https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.data"
#load data into rR
raw_breast_cancer<-read.csv(data_url,header = F,
                            col.names = names(BreastCancer))
# This is done to keep a copy of original data
formatted_breast_cancer<-raw_breast_cancer
#create a vector dict used to change values to character
dict<-c("2"="benign","4"="malignant")
#functions coerces values of dict into characters
map_class<-function(x)dict[as.character(x)]
formatted_breast_cancer$Class<-formatted_breast_cancer$Class %>% 
  map_class %>% 
  unname %>% 
  factor(levels = c("benign","malignant"))
str(formatted_breast_cancer)
```

$formatted\_breast\_cancer\$Class<-formatted\_breast\_cancer\$Class$ is used to overwrite current values with the new values defined on the RHS of the $<-$ operator
The code below perfomes the same task as the one above but is more user friendly
```{r}
#delete the formatted_breast_cancer and dict  variables to avoid confusing R
rm(formatted_breast_cancer)
rm(dict)
# Copy original data to another variable
formatted_breast_cancer<-raw_breast_cancer
# create an empty vector of characters of the same size as the Class variable
dict<-vector(mode="character",length = length(formatted_breast_cancer$Class))
# changing numerics to characters
dict[formatted_breast_cancer$Class==2]<-"bening"
dict[formatted_breast_cancer$Class==4]<-"malignant"
#editing existing Class variable to accommodate new data that is a factor
dict
formatted_breast_cancer$Class<-factor(dict,levels=c("benign","malignant"))
str(formatted_breast_cancer)

```


```{r}
raw_breast_cancer$Class %>% 
  {dict<-c("2"="benign","4"="malignant")
  dict[as.character(.)]
  } %>% 
  unname %>% 
  factor(levels= c("benign","malignant")) %>% 
  table
```

-   The code chunk above defines all we've done
-   Try fixing the latter code to get similar results.
-   The next save is to save our imported data to avoid parsing (ie,
    avoiding repeating the whole process
-   `save()` function is used to save our new file
-   we'll save the file in the same directory we created before

```{r}
formatted_breast_cancer %>% 
save(file="C:/Users/ganje/OneDrive/Documents/project/R_work/data/formatted-breas-cancer.rda")
```

*NB:* use `/` for your directory's pathway as R interprets `\` as escape
character

The suffix `.rda` states file is an R document

The `load` function is then used to load the data set we've saved into
R's `Gloabal enviroment`

-   Thus there's no need to assign it a new variable as it will be
    loaded with its saved variable

```{r}
load("C:/Users/ganje/OneDrive/Documents/project/R_work/data/formatted-breas-cancer.rda")
formatted_breast_cancer %>% head(3)
```

2.  **Boston Housing Data set**

-   It's a data set from *`mlbench`* package and has information about
    crime rates and some explanatory variables that can be used to
    predict crime rates.

```{r}
  library('mlbench')
  data(BostonHousing)
  str(BostonHousing)
```

-   As before we'll load the data directly from its URL(website)
-   [*Click here*](http://tinyurl.com/zq2u8vx) to open the *`tinyurl`*
    of the website to the actual data
    -   **To get the actual link:**
        1.  Open your web browser.
        2.  Copy and paste the `TinyURL` <http://tinyurl.com/zq2u8vx>
            into the address bar.
        3.  On the `TinyURL` website, you'll automatically be redirected
            to the original URL
        4.  The actual link associated with the
            `TinyURL`"<https://archive.ics.uci.edu/ml/machine-learning-databases/housing/housing.data>"
            will be displayed in your browser's address bar.
-   We'll save the original URL in a variable, `data_url`

```{r,eval=F}
    data_url<-"https://archive.ics.uci.edu/ml/machine-learning-databases/housing/housing.data"
    boston_housing<-read.table(data_url)
    str(boston_housing)
```

-   Boston data set is separated by white spaces thus use of
    read.table()function
-   If we compare the `str(BostonHousing)`(Data set *from`mlbench`*) and
    `str(boston_housing) (imported data set)`, we see that their
    observations are same.
-   Our imported data has different types in each column unlike the
    `BostonHousing` which has all its columns as "numeric" except the
    *`Chas`* variable which is a "factor
-   This can be fixed using the `colClasses` argument.
-   To do this will make a vector of strings for the classes,\
    Vector should be numeric for all columns other than the `chas`
    variable

```{r}
col_classes<-rep("numeric",length(BostonHousing))
col_classes[which("chas" == names(BostonHousing))]<-"factor"
col_classes
```

```         
 - The code above creates a vector of strings.

   - `rep()` function replicates the string "numeric" 14 times ie(the length of the `BostonHousing` dataset).
    - The strings(characters) are stored in a vector of strings, `col_classes`
    _ The `which()` is a boolean function(ie; Displays indices that satisfy the condition as `TRUE`) ie;
```

```{r}
          which("chas" == names(BostonHousing))
```

```         
    - It's used in our code to access the column in the `BostonHousing` that's named `chas`
    - Once accessed the string "factor" is saved as the new data type of that column.
    - The code can also be evaluated as
```

```{r}
        col_classes<-rep("numeric",length(BostonHousing))
        col_classes[4]<-"factor"
        col_classes
```

```         
- Next is to name the columns of our imported dataset
  * The `col.names()` is used to do so
  * We'll equate it to `names(BostonHousing)` to get its column names as the names of our imported data set rather than writing each of them.
- Also, we'll specify the data types of our columns using the `colClasses`, which will be equated to our vector of strings `col_classes` that defines these datatypes,
  
```

```{r,eval=F}
      boston_housing<-read.table(data_url,
      col.names=names(BostonHousing),
      colClasses=col_classes)
      str(boston_housing)
```

-   The imported data can also be manipulated as follows:

```{r}
raw_data<-read.table(data_url,col.names = names(BostonHousing))
boston_housing<-raw_data %>% 
  mutate(across(!chas,as.numeric)) %>% 
  mutate(chas=factor(chas,level=0:1))
str(boston_housing)
```

# Manipulating Data with `dplyr`

[***Click here***](https://dplyr.tidyverse.org/articles/dplyr.html) **to
open documentation of dplyr**\
We'll start by installing and loading `dplyr` in R

```{r,eval=F}
install.packages('dplyr')
```

```{r, warning=F}
library('dplyr')
```

## Some useful functions of `dplyr`

1.  **`tbl_df()`**

    -   its used to convert a `dataframe` into a `tibble`
    -   A tibble is a modern modified `dataframe` that has more features
        than the `dataframe`

    ```{r,warning=F, comment=NA}
      iris%>%tbl_df
      iris %>% head(10)
    ```

-   It only prints the 1^st^ 10 rows and not all columns are printed
-   Compare `iris` data set as a `tibble` and as a `dataframe` in the
    out put of our code

2.  **`select()`**\

-   Picks selected columns and gets rid of the rest. A. **Picking 1
    column**

    -   Its equivalent to selecting columns in `dataframes` using
        indexing

    ```{r,warning=F,comments=NA}
      iris %>% tbl_df %>% select(Petal.Width) %>% head(3)
    ```

-   In the above code:

    -   Our data set `iris` is transformed to a `tibble`

    -   We select the `Petal.width` column using `select()`

        -   *We'll repeat the same code using indices*

    -   Using the `head()` we select the 1^st^ 3 rows

    -   code can also be written as:

        -   *A. Using indices*

        ```{r,warning=F,comment=NA}
            iris %>%tbl_df %>%.[,4]%>% head(3)
        ```

        -   *B. Using column name*

        ```{r,warning=F,comment=NA}
            iris %>% tbl_df %>% .$Petal.Width %>% head(3)
        ```

    B. **Picking several columns**

    ```{r,warning=F,comment=NA}
        iris %>% tbl_df %>% 
          select(Sepal.Width,Petal.Length) %>% head(3)
    ```

    C. **Picking a range of columns**

    ```{r,warning=F,comment=NA}
        iris %>% tbl_df %>% 
          select(Sepal.Length:Petal.Length) %>% head(3)
    ```

    -   **`select()`'s effective use is in pattern matching when
        selecting columns**

    -   This enables picking columns based on column names.

    -   To do this, we have some specified arguments based on what you
        want -Below are examples with their specified arguments

        -   Example_1 ("starts_with"):

        ```{r,warning=F,comment=NA}
              iris %>% tbl_df %>%
              select(starts_with("Petal")) %>% head(3)
        ```

        ```         
          Columns whose names starts with "Petal" are printed
        ```

        -   Example_2 ("ends_with")

        ```{r,warning=F,comment=NA}
              iris %>% tbl_df %>% 
              select(ends_with("Width")) %>% head(3)
        ```

        ```         
           Columns whose names end with "Width" are printed
        ```

        -   Example_3 ("contains")

        ```{r,warning=F,comment=NA}
              iris %>% tbl_df %>% 
              select(contains("etal")) %>% head(3)
        ```

        ```         
           Columns whose names contain "etal" are printed
        ```

        -   Example_4 ("matches")

        ```{r,warning=F,comment=NA}
              iris %>% tbl_df %>% 
              select(matches("t")) %>% head(3)
        ```

        ```         
           Columns whose names cantain the string "t" are printed
           **NB:** Columns whose names start or end with "t" are not printed
        ```

        -   `select()` can be used to remove columns

        ```{r,warning=F,comment=NA}
              iris %>% tbl_df %>% 
              select(-starts_with("Petal")) %>% head(3)
        ```

3.  **`mutate()`**

-   Adds computed values to your `dataframe`
-   Allows adding of a column to a `dataframe` by specifying the
    expression to compute it(ie; expression to compute column to be
    added)

```{r,warning=F,comment=NA}
  iris %>% tbl_df %>% 
  mutate(P_W_plus_P_L=Petal.Width+Petal.Length) %>% 
  select(Species,P_W_plus_P_L) %>% head(3)
```

-   `P_W (Petal.width), P_L (Petal.Length)`
-   More columns can be added by specifying them in `mutate()`

```{r,warning=FALSE,comment=NA}
iris %>% tbl_df %>% 
  mutate(P_W_plus_P_L=Petal.Width+Petal.Length,
         S_W_plus_S_L=Sepal.Width+Sepal.Length) %>% 
  select(P_W_plus_P_L,S_W_plus_S_L) %>% head(3)
```

-   `S_W (Sepal.Width), S_L (Sepal.Length)`
-   `mutate()` can be called multiple times in the pipeline

4.  **`transmute()`**

-   Modification of the `select()` and `mutate()`
-   Just like `mutate()` it adds computed values of a column to a data
    frame and just like `select()` it picks columns whose values are
    computed and get rids of the rest.

```{r,warning=F,comment=NA}
  iris %>% tbl_df %>% 
  transmute(P_W_plus_P_L=Petal.Width+Petal.Length) %>% head(3)
```

5.  **`arrange()`**

-   Rearranges `dataframes` by sorting given columns to specifications
    given
-   By default it sorts a given column in ascending order

```{r,warning=F,comment=NA}
  iris %>% tbl_df %>% 
  arrange(Sepal.Length) %>% head(6)
```

-   By using the argument `desc` it sorts given columns in descending
    order

```{r,warning=F,comment=NA}
iris %>% tbl_df %>% 
  arrange(desc(Sepal.Length)) %>% head(6)
```

6.  **`filter()`**

-   Picks out rows based on logical expressions

```{r,warning=F,comment=NA}
iris %>% tbl_df %>% 
  filter(Sepal.Length > 5) %>% head(3)
```

-   You can have at least 1 logical expressions

```{r,warning=F,comment=NA}
iris %>% tbl_df %>% 
  filter(Sepal.Length >5,Species=="virginica") %>% 
  select(Species,Sepal.Length) %>% head(3)
```

7.  **`group_by()`**

-   It splits data into sub tables based on column values
-   Tells R you want to work with data in different subsets\`
-   You can group data by 1 or more variables
-   Works together with `summarize()` or `summarise()`
-   Is effective when grouping factors or discrete numbers11

```{r,warning=FALSE,comment=NA}
  iris %>% tbl_df %>% group_by(Species)
```

-   There's no much difference here when using `group_by()` as the
    output is similar to original data set

8.  **`summarise/summarize()`**

-   Calculates summary of descriptive statistics

-   Let's you compute different descriptive statistics by expressing
    what you want to summarize

-   Like's of descriptive statistics include:

    a.  `mean` - Central tendancy
    b.  `variance` - Deviation from mean (Mean Square Error)
    c.  `sd` - Square root of mean Square Error
    d.  `covariance`
    e.  `cor-relation`
    f.  `mode`
    g.  `median`
    h.  `skewness`
    i.  `kurtosis`
    j.  `range`
    k.  `Interquartile Range (IQR)`
    l.  `sample and population size`

    ```{r}
      iris %>% 
        summarise(mean(Petal.Length),
      mean(Sepal.Length))
    iris %>% summarize(sd(Petal.Length),
                       sd(Sepal.Length))
    iris %>% 
    summarise(mean(Petal.Length),
      mean(Sepal.Length),sd(Petal.Length),
                       sd(Sepal.Length))
    ```

    ```{r}
    iris %>% 
      summarise(observations=n())
    ```

9.  **`group_by()` and `summarise()`**

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(mean_P_L=mean(Petal.Length),
            mean_S_L=mean(Sepal.Length))
```

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(Number_of_species=n())
```

-   `summarise()` function can take at least 1 argument which are
    descriptive statistics

```{r}
  iris %>% 
  group_by(Species) %>% 
  summarise(Number_of_obser=n(),
            mean_PetalLength=mean(Petal.Length),
            var_PetalLength=var(Petal.Length),
            sd_PetalLength=sd(Petal.Length))
```

# Breast Cancer Data Manipulation with `dplyr`

1.  Using `mutate()` to express `class` as a factor

```{r}
formatted_breast_cancer<-raw_breast_cancer %>% mutate(Class=Class %>% {
  c("2"="benign","4"="malignant")[as.character(.)]
} %>% 
  unname %>% 
  factor(levels= c("benign","malignant"))
)
formatted_breast_cancer %>% head(3)
str(formatted_breast_cancer)
```

```{r}
formatted_breast_cancer<-raw_breast_cancer %>% mutate(Class=Class %>% {
  c("2"="benign","4"="malignant")[as.character(.)]
} %>% 
  unname %>% 
  factor(levels= c("benign","malignant"))
) %>% 
  mutate(Id=as.character(Id)) %>% 
  mutate(across(Cl.thickness:Epith.c.size,as.ordered)) %>% 
  mutate(across(Bare.nuclei:Mitoses,as.factor)) %>% 
  mutate(Bare.nuclei=factor(Bare.nuclei,levels=1:10))
  str(formatted_breast_cancer)


```

-   **NB:** In the chunk above `mutate()` doesn't add a new column
    `Class` but writes over the existing column `Class` as they share
    the same variable name

-   If keen, we've not used the dictionary table `dict` as we can't have
    it inside the mutate function

-   if it was to be used, we have to define it 1^st^ as a function
    outside `mutate()`

```{r}
format_class<-. %>% {
  dict<-c("2"="benign","4"="malignant")
  dict[as.character(.)]
} %>% unname %>% factor(levels=c("benign","malignant"))
formatted_breast_cancer<-raw_breast_cancer %>% mutate(Class=format_class(Class))
```

2.**Using `group_by()`and `summarise()`**

-   We want to check how different variables affect the response
    variable `Class`

```{r}
formatted_breast_cancer %>% 
  group_by(Class) %>% 
  summarise(mean.thickness=mean(Cl.thickness))
```

```{r}
formatted_breast_cancer %>% 
  group_by(Class) %>% 
  summarise(mean.size=mean(Cell.size))
```

3.  **Using `group_by()`, `summarise()`,`arrange()`**

-   We want to check number of occurrence of `benign` and `malignant`
    tumor as cell size increase

```{r,comment=NA}
  formatted_breast_cancer %>% 
  arrange(Cell.size) %>% 
  group_by(Cell.size,Class) %>% 
  summarise(class_count=n())
```

We learn:

-   There are more `benign` tumors compared to `malignant` when cell
    size is small.
-   There are more `malignat` tumors than `benign` when cell size is
    large
-   Occurrence of `benign` is inversely related to increase in cell size
-   Occurrence of `malignant` is directly related ti increase in cell
    size

Let's see how cell thickness behaves as a function of both class and
cell size

```{r}
formatted_breast_cancer %>% 
  group_by(Class,as.factor(Cell.size)) %>% 
  summarize(mean.thickness=mean(Cl.thickness))
```

We learn that:

-   `mean.thickness` increases as cell size increases for `benign`
    whilst there's no pattern for the `malignant` tumor

Let"s look at the number of `benign` and `malignant` tumors for each
cell size and see what the thickness is

```{r}
formatted_breast_cancer %>% 
  group_by(as.factor(Cell.size),Class) %>% 
  summarize(mean.thickness=mean(Cl.thickness))
```

## Tidying data with `tidyr`

[\*CLICK HERE] (\*<https://tidyr.tidyverse.org/articles/tidy-data.html>)
to access package's documentation

our 1^st^ step is to install and load the package

```{r,eval=F}
install.packages("tidyr")
```

```{r,warning=F,comment=NA}
library("tidyr")
```

Tidying data helps in plotting and summarizing data efficiently

Tidying data will help one work with columns of a data frame you want to
work with

For instance in the `iris` data set we can check how `Petal.Width`
varies with different `Species`

-   We can use the `dplyr` function `select()` to check it out

```{r}
iris %>% select(Species,Petal.Length) %>% head(3)


```

We've said that tidying data enables us specify columns we want to work
with

This makes it efficient for us to summarize and plot data

Let's say we want a `boxplot`of `Petal.Length` against `Specie` of the
`iris` data set

Steps:

-   Specify the columns we want to work with

-   Use `qplot()` of `ggplot` package for plotting

```{r}
iris %>% select(Species,Petal.Length) %>% 
  qplot(Species,Petal.Length,data=.,geom="boxplot")

```

```{r}
iris %>% 
  ggplot(aes(x=Species,y=Petal.Length)) +
  geom_boxplot()
```

Our plotting code works since we have columns for the x and y axis

In the event we want a plot for at least 2 columns we employ the `tidyr`
package as its difficult to map various different columns in the x and y
axis

We employ the `gather()`

The `gather()` is used to collect columns of a data set

It has 3 arguments:

-   first argument is a column that stores as factors

-   second argument stores values of chosen columns
