---
title: "Regression Analysis Assignment 1"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

| Name      |             RegNo |
|:----------|------------------:|
| 1\. David | S030-01-1925/2022 |
| 2\. Ken   | S030-01-1916/2022 |
| 3\. Lucy  | S030-01-19xx/2022 |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = F)
```

```{r}
library("ggplot2")
library("magrittr")
```

# Question 1

## Solution for (i), (ii)

```{r}
#Age(years)
x<-c(18,23,25,35,65,54,34,56,72,19,23,42,18,39,37)
#max rate
y<-c(202,186,187,180,156,169,174,172,153,199,193,174,198,183,178)
#A plot of the scatter plot
plot(x,y,
     xlab = 'Age(years)',
     ylab = 'Max Rate',
     main = 'Plot of Max Rate on Age')
#create a simple linear model
fit_model<-lm(y~x)

#plotting fitted line
abline(lm(fit_model))
```

## Solution (iii), (iv)

We test the hypothesis: $H_0:{\beta_1}=-1\ vs\ H_1:{\beta_1}\ne -1$

Test statistics is given by:

$t=\frac{\hat{\beta_1}-\beta_1}{\mathcal{SE(\hat{\beta_1})}}$

```{r}
#define slope for null hypothesis
beta1<- -1
#Get the fitted slope
beta1_hat<-round(fit_model$coefficients[2],4)
#store summary of model as an object to access standard error of estimates
summ_model<-summary(fit_model)
#Get standard error of slope
se_beta1_hat<-round(summ_model$coefficients[2,"Std. Error"],5)
#Get test statistics
t_test<-round((beta1_hat-beta1)/se_beta1_hat,4)
cat("Computed test statistics is given by :", t_test)
#Get p-value
(pval<-2*pt(t_test,13,lower.tail = F))
cat("P-value is given by:", pval)

```

# ebu angalia io p-value tena

# Question 2

## Solution (i)

```{r}
load("smoking_data.RData")
smoking_data<-data.frame(country,alcohol,cigarettes)
smoking_data %>% ggplot(aes(x=alcohol,y=cigarettes,col=country))+
  geom_point()+
  geom_smooth(method = "lm",fill=NA)+
  geom_rug()+
  xlab("Alcohol Consumption")+
  ylab("Cigarettes Consumption")+
  ggtitle("ScatterPlot of Cigarettes vs Alcohol consumption of various Countries")+
  theme(plot.title = element_text(size = 12, face="bold"))
  
```

There exist a linear relationship between the data on alcohol and cigarette consumption, where increase in alcohol consumption increase cigarette consumption.

## Solution (ii)

```{r}
#Pearson Correlation coefficient
p_cor<-cor(alcohol,cigarettes,method="pearson")
cat("The pearson's correlation coefficient is:",round(p_cor,4))
```

## Solution (iii)

```{r}
alcohol.2<-alcohol[-c(6,16)]
cigarettes.2<-cigarettes[-c(6,16)]

```

The above code removes elements in the 6^th^ and 16^th^ indices(positions) respectively for both the alcohol and smoke variables and stores elements of the remaining indices in the variables alcohol.2 and cigarettes.2 respectively.

The code is used to delete elements whose indices are specified.

## Solution (iv)

```{r}
plot(alcohol.2,cigarettes.2,
     xlab="Alcohol Consumption",
     ylab ="Cigarette Consumption")
title("Scatter Plot of Cigarettes vs Alcohol of New Data")

```

```{r}
#Adjusted pearson coreelation coefficient on alcohol and cigarettes
Ap_cor<-cor(alcohol.2,cigarettes.2,method = "pearson")
cat("The Adjusted pearson correlation coefficient based on new data is:", round(Ap_cor,4))

```

There's a strong positive correlation between alcohol and cigarette consumption in the adjusted data than there is in the original data.

## Solution (iv)

We test the hypothesis $H_0:\rho=0\ vs\ H_1:\rho\ne0$

Test statistics is given by $t=\frac{r\sqrt{n-2}}{\sqrt{1-r^2}}\sim \mathcal{t(n-2)}$

```{r}
#get correlation
r<-round(Ap_cor,4)
#Check equality in sie of variables
length(cigarettes.2)==length(alcohol.2)
#get size of variables
n=length(alcohol.2)
#Get test statistics
(t_test<-(r*sqrt(n-2))/sqrt(1-r^2))
cat("Computed test statistics is:", round(t_test,4))
#get p-value
(pval<-2*pt(t_test,12,lower.tail = F))
cat("P-value is given by:",pval)
```
